Compositional Equivalences Based on Open pNets

### 2022-09-27 2016-Weak-FH-Bisim/WeakBisim.tex
Submission weak bisim + work on the refinement paper (body)
advancing on remarks

\documentclass{elsarticle}
%\overfullrule=1mm
\usepackage[multiple]{footmisc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

@@

\title{Compositional equivalences based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\ead{rabea.ameur-boulifa@telecom-paris.fr}

\author{Ludovic Henrio}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\ead{ludovic.henrio@cnrs.fr}

@@

\begin{abstract}

@@

pNet is a generalisation of automata that model open systems. They feature variables and hierarchical composition. Open pNets are pNets with holes, i.e. placeholders inside the hierarchical structure  that can be filled later by sub-systems. 
However, there is no standard tool for defining the semantics of an open system in this context, this article first defines \emph{open automata} that are labelled transition systems with parameters and holes.

Relying on open automata, this article defines bisimulation relations  for the comparison of systems specified as pNets.
 We first define a strong bisimulation for open pNets.
  We then define an equivalence relation similar to
  the classical \emph{weak bisimulation}, and study its properties.
Among these properties we are interested in compositionality: if two systems are proven equivalent they will be undistinguishable by their context, and they will also be undistinguishable when their holes are filled with equivalent systems.
We identify sufficient conditions to ensure compositionality of strong and weak bisimulation.
The article is illustrated with a transport protocol running example.
\end{abstract}

@@

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
These works use the term \emph{parameter} to designate systems where variables that have a strong influence the system structure and behaviour. In parameterised systems, parameters can typically be the number of processes in the system or the way they interact.
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. 30 years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

This article provides a theoretical background that allows us to implement such a verification 
platform. We build upon the concept of pNets that allowed us to give a behavioural semantics 
of distributed components and verify the correctness of distributed applications in the past 15 years. pNets  is a
low level semantic framework for expressing the behaviour of various
classes of distributed languages, and as a common internal format for
our tools.  pNets allow the
specification of parameterised hierarchical labelled
transition systems:  labelled transition systems with parameters can be
combined hierarchically.

We develop here a semantics for a model of interacting processes with parameters and holes. 
Our approach is originally inspired from Structured Operational Semantics with conditional premisses as in \cite{GROOTE1993263,VANGLABBEEK2004229}.
But we aim at a more constructive and implementable approach to compute the semantics (intuitively transitions including first order predicates) and to 
check equivalences for these open systems.
The main interest of our symbolic approach is to define a method to
prove properties directly on open structures; these properties will then be preserved
by any correct instantiation of the holes.
As a consequence, our model allows us to reason on composition
operators as well as on full-size distributed systems. The parametric
nature of the model and the properties of compositionability of the
equivalence relations are thus  the main
strengths of  our approach.

@@

\paragraph{A semantics for Open pNets based on open automata}
The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no hole. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion.

@@

\subsubsection*{Previous Works and Contribution}
While most of our previous works relied on closed, fully-instantiated semantics~\cite{BBCHM:article2008,AmeurBoulifa2017,HKM-FASE16}, it is only recently that we could design a first version of a  parameterised semantics for pNets with a strong bisimulation equivalence~\cite{henrio:Forte2016}. This article builds upon this previous parameterised semantics and provides a clean and complete version of the semantics with a slightly simplified formalism that makes proofs easier. It also adds a notion of global state to automata.  Also, in \cite{henrio:Forte2016} the study of compositionability was only partial, and in particular the proof that bisimulation is an equivalence is one new contribution of this article and provides a particularly interesting insight on the semantic model we use.
The new formalism allowed us to extend the work and define weak bisimulation  for open automata, which is entirely new. This allows us to define a weak bisimulation equivalence for open pNets with valuable properties of compositionality. 
To summarise, the contribution of this paper are the following:
\begin{itemize}
\item The definition of open automata: an algebra of parameterised automata with holes, and a strong bisimulation equivalence. This is an adaptation of~\cite{henrio:Forte2016} with an additional property stating that strong bisimulation equivalence is indeed an equivalence relation.
\item A semantics for open pNets expressed as translation to open automata. This is an adaptation of~\cite{henrio:Forte2016} with a complete proof that strong bisimulation is compositional.
\item A theory of weak bisimulation for open automata, and its properties. It relies on the definition of weak open transitions that are derived from transitions of the open automaton by concatenating invisible action transitions with one (visible or not) action transition. The precise and sound definition of the concatenation is also a major contribution of this article.
\item A resulting weak bisimulation equivalence for open pNets and a simple static condition on synchronisation vectors inside pNets that is sufficient to ensure that weak bisimulation is compositional.
\item An illustrative example based on a simple transport protocol, showing the construction of the weak open transitions, and the proof of weak bisimulation.
\end{itemize}

@@

\section{Background and notations}\label{sec:notations}
This section introduces the notations we will use in this article, and  recalls the definition of pNets~\cite{henrio:Forte2016} with an informal semantics  of the pNet constructs. The only significant difference compared to our previous definitions is that we remove here the restriction that was stating that variables should be local to a state of a labelled transition system.

@@

@@

%\LUDO{introduce magic op}
%$\CreateISet{a}{i}$ creates a single entry indexed set $(i\mapsto a)$ \emph{but only if 
%$a$ is not already an indexed set.}

@@

\begin{definition}[Sorts, Holes, Leaves, Variables of pNets]\label{def-sortholeleave}~~

@@

pNets have been used to model GCM distributed component systems, illustrating the expressiveness of the model~\cite{AmeurBoulifa2017}. 
These works show that pNets are convenient to express the behaviour of a system in a compositional way. 
Unfortunately, the semantics of pNets and the existing tools at this point were only able to deal with a closed system completely instantiated: pNets could be used as composition operator in the definition of the semantics, which was sufficient to perform finite-state model checking on a closed system, but there was no theory for the use of pNets as operators and no tool for proving properties on open system. 
Consequently, much of the formalisation efforts did not use holes and the interplay between holes, sorts, and synchronisation vector was not formalised.
In previous works~\cite{AmeurBoulifa2017}, only closed pNets were equipped with a semantics, it was defined as labelled transition systems. 
%The old semantics is not necessary to understand this article but it defines the semantics of pNets in a simpler and more operational setting. 
The theory of pNets as operators able to fully take into account open systems is given in the following sections. Comparing formally the existing direct operational semantics and the semantics derived from open automata in the current article would be an interesting partial proof of soundness for our semantics. The proof could only be partial as the formal semantics that exists only consider closed and fully instantiated pNets. Proving 
% a bisimulation
an equivalence between the semantics presented in this article and the operational one shown in~\cite{AmeurBoulifa2017} is outside the scope of this article because we focus here on the modelisation of holes that were not modelled in the previous semantics.

\subsection{Running Example}
To illustrate this work, we use a simple communication protocol, that provides safe transport of data between two processes, over unsafe media.

@@

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on a set of variables (potentially with only symbolic values). 
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$, a set $V$ of variables, and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$ is the set of holes involved in the transition; $s, s'\in\mathcal{S}$ are states of the automaton; and $\beta_j$
        is a transition of the hole $j$, with $\Sortop(\beta_j)\in\Sort_j$. $\alpha$  is the resulting action of this open transition.
%        We call \emph{source variables} of an OT the union of all variables
%        in the different terms
%        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate, \Post\ is a set of 
	assignments that are effective after the open transition, they are
        represented as a substitution function: $({x_k\gets e_k})^{k\in K}$.
Predicates and expressions of an open transition can refer to the variables in $V$, and
        in the different terms
        $\beta_j$ and $\alpha$. More precisely:

@@

It is important to understand the difference between the red dotted rule and a classical 
inference rule. They correspond to two different logical levels.
On one side, classical (black) inference rules  use  an expressive logic (like any other computer science article).
 On the other side, open transition rules (with dotted lines) are logical implications, but using a  logic with a specific syntax and that can be mechanized (this logic includes the boolean expressions $\AlgB$, boolean operators, and term equality).
%  Open transitions could typically be handled in a 
% mechanised way.

@@

\begin{thm}[FH-Bisimulation is an equivalence]\label{thm-equiv} Suppose $\mathcal{R}$ 
is an FH-bisimulation. Then $\mathcal{R}$ is an equivalence, that is, $\mathcal{R}$ is 
reflexive, symmetric and transitive.
\end{thm}

@@

\begin{example}[CCS choice (counter-example)]\label{ex:CCSplus}
Here is the encoding of a choice operator.\\[1.5ex]
\begin{tabular}{cp{.53\linewidth}}
 \begin{minipage}{.4\linewidth}
 \includegraphics[width=1.04\linewidth]{XFIG/Choice.pdf}\vspace{-18ex}
\end{minipage} 
&
The left hole is
  indexed $l$ the right hole $r$. The third subnet, contains an LTS encoding
  the control part. We obtain the specific behaviour with the 
  synchronisation vector. The first action  of one of
  the holes decides which branch of the LTS is activated;
   all subsequent actions will be from the same side. 
\\
\end{tabular}

This pNet does not satisfy Definition~\ref{def:non-obspNet}. Indeed, if $a$ or $b$ is $\tau$ then the $+$ operator can indeed observe the $\tau$ transition.
On the other side, the parallel operator defined similarly \emph{satisfies} Definition~\ref{def:non-obspNet}.

@@

\begin{example}[CCS Choice] Consider the $+$ operator of CCS, shown in  Example~\ref{ex:CCSplus}.  It is well-known that weak bisimulation is not a congruence in CCS, and this is reflected here because we have shown that the $+$ operator can observe the $\tau$ transitions. Thus, even if we can define a weak bisimulation for CCS with $+$ it does not verify the necessary requirements for being a congruence.
\end{example}

@@

\subsubsection*{Open and Compositional systems}
%\TODO{retravailler le paragraphe ci dessous: faire ref a Facs-2013 (reconfig ds pNets) - virer 14 (probabilistic) ?} \cite{GasparHM:2013}
In
\cite{Johnson:2013:CBSE,Johnson:2014:qosa}, the authors investigate several methodologies for the compositional verification of software systems, with the aim to verify reconfigurable component systems. To improve scaling and compositionality, the authors  decouple the verification problem that is to be resolved by a SMT (satisfiability modulo theory) solver into independent sub-problems on independent sets of variables. These works clearly highlight the interest of incremental and compositional verification in a very general setting. In our own work on open pNets, adding more structure to the composition model, we show how to enforce a compositional proof system that is more powerful than independent sets of variables. Our theory has also been encoded into an SMT solver and it would be interesting to investigate how the examples of evolving systems studied by the authors could be encoded into pNet and verified by our framework. However, the models of Johnson et al. are quite different from ours, in particular they are much less structured, and translating them is clearly outside the scope of this article.
In previous work \cite{gaspar:hal-00916115}, we also have shown how (closed) pNet models could be used to encode and verify finite instances of reconfigurable component systems. 
%Additionally, the more compositional constraint resolution provided by the authors could also be interesting to make our SMT proofs scale better (inside a single pNet for example), the integration of the Z3 framework described in \cite{Johnson:2014:qosa} with our proof theory for open pNets would also be worth investigating from a practical point of view. 
%One major different however is that pNets in their current state are not much adapted to probabilistic verification contrarily to \cite{Johnson:2014:qosa} but they are better suited to verify complex behavioural properties.

@@

\section{Conclusion and Discussion}
\label{section:conclusion}
 pNets (Parameterised Networks of Automata) is a formalism adapted to the representation of the behaviour of a parallel or distributed systems. One  strength of pNets is their parameterised nature, making them adapted to the representation of systems of arbitrary size, and making the modelling of parameterised system possible. Parameters are also crucial to reason on interaction protocols that can address one entity inside an indexed set of processes. pNets have been successfully used to represent behavioural specification of parallel and distributed components and verify their correctness~\cite{AmeurBoulifa2017,HKM-FASE16}. VCE is the specification and verification platform that uses pNets as an intermediate representation.

>>>>

\documentclass{elsarticle}
%\overfullrule=1mm
\usepackage[multiple]{footmisc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}

\lstset{basicstyle=\small}

@@

\title{Compositional Equivalences Based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\ead{rabea.ameur-boulifa@telecom-paris.fr}

\author{Ludovic Henrio\corref{cor1}}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\ead{ludovic.henrio@cnrs.fr}

@@

\cortext[cor1]{Corresponding author}
\begin{abstract}

@@

pNet is a generalisation of automata that model open systems. They feature variables and hierarchical composition. Open pNets are pNets with holes, i.e. placeholders inside the hierarchical structure  that can be filled later by sub-systems. 
However, there is no standard tool for defining the semantics of an open system in this context. This article first defines \emph{open automata} that are labelled transition systems with parameters and holes.
Relying on open automata, it then defines bisimulation relations  for the comparison of systems specified as pNets.
 We first present a strong bisimulation for open pNets.
Next we offer an equivalence relation similar to
  the classical \emph{weak bisimulation} equivalence, and study its properties.
Among these properties we are interested in compositionality: if two systems are proven equivalent they will be undistinguishable by their context, and they will also be indistinguishable when their holes are filled with equivalent systems.
We identify sufficient conditions to ensure compositionality of strong and weak bisimulation.
The contributions of this article are illustrated using a transport protocol as running example.
\end{abstract}

\begin{keyword}
Bisimulation, compositionality, automata, semantics.
\end{keyword}

@@

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
These works use the term \emph{parameter} to designate  variables whose value have a strong influence the system structure and behaviour. In parameterised systems, parameters can typically be the number of processes in the system or the way they interact.
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. Thirty years later, no
practical verification method and no verification platform use this kind of approaches to provide proof methods for value-passing processes or open process expressions.

This article provides a theoretical background that allows us to implement such a verification 
platform. We build upon the concept of pNets that we have employed to give a behavioural semantics 
of distributed components and verify the correctness of distributed applications in the past 15 years. pNets  is a
low level semantic framework for expressing the behaviour of various
classes of distributed languages, and as a common internal format for
our tools.  pNets support the
specification of parameterised hierarchical labelled
transition systems:  labelled transition systems with parameters can be
combined hierarchically.

We develop here a semantics for a model of interacting processes with parameters and holes. 
Our approach is originally inspired from Structured Operational Semantics with conditional premisses as in \cite{GROOTE1993263,VANGLABBEEK2004229}.
But we aim at a more constructive and implementable approach to compute the semantics (intuitively transitions including first order predicates) and to 
check equivalences for these open systems.
The main interest of our symbolic approach is to define a method to
prove properties directly on open structures; these properties will then be preserved
by any correct instantiation of the holes.
As a consequence, our model allows us to reason about composition
operators as well as about realistic distributed systems. The parametric
nature of the model and the properties of compositionality of the
equivalence relations are thus  the main
strengths of  our approach.

@@

\paragraph{A semantics for open pNets based on open automata}
The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no hole. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion.

@@

\subsubsection*{Previous works and contribution}
While most of our previous works relied on closed, fully-instantiated semantics~\cite{BBCHM:article2008,AmeurBoulifa2017,HKM-FASE16}, it is only recently that we could design a first version of a  parameterised semantics for pNets with a strong bisimulation equivalence~\cite{henrio:Forte2016}. This article builds upon this previous parameterised semantics and provides a clean and complete version of the semantics with a slightly simplified formalism that makes proofs easier. It also adds a notion of global state to automata.  Also, in \cite{henrio:Forte2016} the study of compositionability was only partial, and in particular the proof that bisimulation is an equivalence is one new contribution of this article and provides a particularly interesting insight on the semantic model we use.
The new formalism allowed us to extend the work and define weak bisimulation  for open automata, which is entirely new. This allows us to define a weak bisimulation equivalence for open pNets with valuable properties of compositionality. 
To summarise, the contribution of this paper are the following:
\begin{itemize}
\item The definition of open automata: an algebra of parameterised automata with holes, and a strong bisimulation equivalence. This is an adaptation of~\cite{henrio:Forte2016} with an additional property stating that strong bisimulation equivalence is indeed an equivalence relation.
\item A semantics for open pNets expressed as translation to open automata. This is an adaptation of~\cite{henrio:Forte2016} with a complete proof that strong bisimulation is compositional.
\item A theory of weak bisimulation for open automata, and its properties. It relies on the definition of weak open transitions that are derived from transitions of the open automaton by concatenating invisible action transitions with one (visible or not) action transition. The precise and sound definition of the concatenation is also a major contribution of this article.
\item A resulting weak bisimulation equivalence for open pNets and a simple static condition on synchronisation vectors inside pNets that is sufficient to ensure that weak bisimulation is compositional.
\item An illustrative example based on a simple transport protocol, showing the construction of the weak open transitions, and the proof of weak bisimulation.
\end{itemize}

@@

\section{Background and Notations}\label{sec:notations}
This section introduces the notations we will use in this article, and  recalls the definition of pNets~\cite{henrio:Forte2016} with an informal semantics  of the pNet constructs. The only significant difference compared to our previous definitions is that we remove here the restriction that was stating that variables should be local to a state of a labelled transition system.

@@

\subsection{The principles of Parameterised Networks (pNets)}
\label{section:pnets-princ}

@@

We introduce the notion of pNets through a simple example below, and define formally pLTSs and pNets afterwards:

\begin{example}[CCS choice]\label{ex:CCSplus}

Here is the encoding of a choice operator.\\[1.5ex]
\noindent\begin{tabular}{cp{.45\linewidth}}
 \begin{minipage}{.49\linewidth}
 \includegraphics[width=1.0\linewidth]{XFIG/Choice.pdf}\vspace{-10ex}
\end{minipage} 
&
 It consists of one pNet with two holes and a sub-net. The left hole is
  indexed $L$ the right hole $R$. The sub-net is an labelled transition system (LTS) with three states and emitting actions $l$ and $r$. 
\\
\end{tabular}

The behaviour of the pNet is defined with 
  synchronisation vectors also shown on the figure. In the examples, we write them on the form $<a,-,l>\to a$. This states that if the first hole $L$ performs the action $a$ and the third sub-net, i.e. the LTS, performs the action $l$, both of them progress synchronously, and an action $a$ is emitted by the pNet. $-$ denotes that the second hole does nothing. On the formal side, numbering and ordering the vectors is cumbersome, this is why we adopt indexed families of actions.
When encoding this operator, the LTS is sometimes called the ``control part'', it controls the evolution of the rest of the pNet. 
The first action  of one of  the holes decides which branch of the LTS is activated;
   all subsequent actions will be from the same side.

\end{example}
%\LUDO{introduce magic op}
%$\CreateISet{a}{i}$ creates a single entry indexed set $(i\mapsto a)$ \emph{but only if 
%$a$ is not already an indexed set.}

\subsection{Parameterised Labelled Transition systems (pLTS)}
\label{section:plts}

@@

\begin{definition}[Sorts, holes, leaves, variables of pNets]\label{def-sortholeleave}~~

@@

pNets have been used to model GCM distributed component systems, illustrating the expressiveness of the model~\cite{AmeurBoulifa2017}. 
These works show that pNets are convenient to express the behaviour of a system in a compositional way. 
Unfortunately, the semantics of pNets and the existing tools at this point were only able to deal with a closed system completely instantiated: pNets could be used as composition operator in the definition of the semantics, which was sufficient to perform finite-state model checking on a closed system, but there was no theory for the use of pNets as operators and no tool for proving properties on open system. 
Consequently, much of the formalisation efforts did not use holes and the interplay between holes, sorts, and synchronisation vector was not formalised.
In previous works~\cite{AmeurBoulifa2017}, only closed pNets were equipped with a semantics, it was defined as labelled transition systems. 
%The old semantics is not necessary to understand this article but it defines the semantics of pNets in a simpler and more operational setting. 
The theory of pNets as operators able to fully take into account open systems is given in the following sections. Comparing formally the existing direct operational semantics and the semantics derived from open automata in the current article would be an interesting partial proof of soundness for our semantics. The proof could only be partial as the formal semantics that exists only consider closed and fully instantiated pNets. Proving 
% a bisimulation
an equivalence between the semantics presented in this article and the operational one shown in~\cite{AmeurBoulifa2017} is outside the scope of this article because we focus here on the modelling of holes that were not considered in the previous semantics. 
\TODO{Check}
It is however easy to see that, in case there is no hole the structure of the open automaton that defines the semantics here is very close to the pLTS that is used in~\cite{AmeurBoulifa2017} to define the semantics even though the formalisms are different.

\subsection{Running example}
To illustrate this work, we use a simple communication protocol, that provides safe transport of data between two processes, over unsafe media.

@@

\subsection{Open automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on a set of variables (potentially with only symbolic values). 
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$, a set $V$ of variables, and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$ is the set of holes involved in the transition; $s, s'\in\mathcal{S}$ are states of the automaton; and $\beta_j$
        is a transition of the hole $j$, with $\Sortop(\beta_j)\in\Sort_j$. $\alpha$  is the resulting action of this open transition.
%        We call \emph{source variables} of an OT the union of all variables
%        in the different terms
%        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate, \Post\ is a set of 
	assignments that are effective after the open transition, they are
        represented as a substitution function: $({x_k\gets e_k})^{k\in K}$.
Predicates and expressions of an open transition can refer to the variables in $V$, and
        in the different terms
        $\beta_j$ and $\alpha$. More precisely:

@@

\TODO{Check this paragraph}
It is important to understand the difference between the red dotted rule and a classical 
inference rule. They correspond to two different logical levels.
On one side, classical (black) inference rules act at the mathematical level of the paper proofs. They use  an expressive logic (like any other computer science article).
 On the other side, open transition rules (with dotted lines) are logical implications that belong to the open automata algebra.
Their  logic has a specific syntax  that can be mechanized; this logic includes the boolean expressions $\AlgB$, boolean operators, and term equality.
%  Open transitions could typically be handled in a 
% mechanised way.

@@

\begin{thm}[FH-bisimulation is an equivalence]\label{thm-equiv} Suppose $\mathcal{R}$ 
is an FH-bisimulation. Then $\mathcal{R}$ is an equivalence, that is, $\mathcal{R}$ is 
reflexive, symmetric and transitive.
\end{thm}

@@

@@

\begin{example}[CCS Choice] Consider the $+$ operator of CCS, shown in  Example~\ref{ex:CCSplus}. 
The pNet does not satisfy Definition~\ref{def:non-obspNet}. Indeed, if $a$ or $b$ is $\tau$ then the $+$ operator can  observe the $\tau$ transition.
It is well-known that weak bisimulation is not a congruence in CCS, this corresponds to the fact that the $+$ operator can observe the $\tau$ transitions. Thus, even if we can define a weak bisimulation for CCS with $+$ it does not verify the necessary requirements for being a congruence.

On the other side, the parallel operator defined similarly \emph{satisfies} Definition~\ref{def:non-obspNet}, and indeed bisimilarity is a congruence for the parallel operator in CCS.
\end{example}

@@

\subsubsection*{Open and compositional systems}
%\TODO{retravailler le paragraphe ci dessous: faire ref a Facs-2013 (reconfig ds pNets) - virer 14 (probabilistic) ?} \cite{GasparHM:2013}
In
\cite{Johnson:2013:CBSE,Johnson:2014:qosa}, the authors investigate several methodologies for the compositional verification of software systems, with the aim to verify reconfigurable component systems. To improve scaling and compositionality, the authors  decouple the verification problem that is to be resolved by a SMT (satisfiability modulo theory) solver into independent sub-problems on independent sets of variables. These works clearly highlight the interest of incremental and compositional verification in a very general setting. In our own work on open pNets, adding more structure to the composition model, we show how to enforce a compositional proof system that is more powerful than independent sets of variables. Our theory has also been encoded into an SMT solver and it would be interesting to investigate how the examples of evolving systems studied by the authors could be encoded into pNet and verified by our framework. However, the models of Johnson et al. are quite different from ours, in particular they are much less structured, and translating them is clearly outside the scope of this article.
In previous work \cite{gaspar:hal-00916115}, we also have shown how (closed) pNet models could be used to encode and verify finite instances of reconfigurable component systems. 
%Additionally, the more compositional constraint resolution provided by the authors could also be interesting to make our SMT proofs scale better (inside a single pNet for example), the integration of the Z3 framework described in \cite{Johnson:2014:qosa} with our proof theory for open pNets would also be worth investigating from a practical point of view. 
%One major different however is that pNets in their current state are not much adapted to probabilistic verification contrarily to \cite{Johnson:2014:qosa} but they are better suited to verify complex behavioural properties.

@@

\section{Conclusion and discussion}
\label{section:conclusion}
 pNets (Parameterised Networks of Automata) is a formalism adapted to the representation of the behaviour of a parallel or distributed systems. One  strength of pNets is their parameterised nature, making them adapted to the representation of systems of arbitrary size, and making the modelling of parameterised system possible. Parameters are also crucial to reason on interaction protocols that can address one entity inside an indexed set of processes. pNets have been successfully used to represent behavioural specification of parallel and distributed components and verify their correctness~\cite{AmeurBoulifa2017,HKM-FASE16}. VCE is the specification and verification platform that uses pNets as an intermediate representation.
